---
title: "Getting started with YLLgmethods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with YLLgmethods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

`YLLgmethods` estimates **years of life lost (YLL)** under the
parametric g-formula with a discrete-time hazard model. This vignette
walks through the typical workflow on the bundled toy dataset.

```{r}
library(YLLgmethods)
data(yll_toy)
str(yll_toy)
```

## ATE: average treatment effect in the whole population

`estimate_yll_gformula_ate()` is a thin wrapper that contrasts
"everyone reference" vs. "everyone exposed" and averages over the
entire study population.

```{r ate, eval = FALSE}
res_ate <- estimate_yll_gformula_ate(
  data = yll_toy,
  B = 50,
  method = "normal",
  show_progress = FALSE,
  id_var = "id",
  time_var = "period",
  event_var = "event",
  exposure_var = "hypertension",
  reference_level = "No",
  exposed_level = "Yes",
  age_at_entry_var = "age",
  age_start = 50,
  age_end = 90,
  age_interval = 5,
  confounders_baseline = c("sex", "education_years", "bmi", "smoke_binary"),
  use_future = FALSE
)

res_ate$summary
```

The result object always contains:

- `summary` — one row per starting age with point estimate, CI, and CI method
- `detailed_results` — point estimates plus per-arm life expectancy and
  both percentile and normal-approximation CIs (when `B > 0`)
- `meta` — call metadata (`estimand`, `B`, `conf_level`, `integration`, …)
- `marginal_survival_point` and `marginal_survival_boot` — marginal
  survival curves used by the `plot_*()` helpers

## ATT and ATC

`_att()` averages the contrast over those who were observed exposed,
`_atc()` over those observed at the reference level. Same arguments,
different target population:

```{r att-atc, eval = FALSE}
res_att <- estimate_yll_gformula_att(
  data = yll_toy,
  B = 50, method = "normal", show_progress = FALSE, use_future = FALSE,
  id_var = "id", time_var = "period", event_var = "event",
  exposure_var = "hypertension",
  reference_level = "No", exposed_level = "Yes",
  age_at_entry_var = "age",
  age_start = 50, age_end = 90, age_interval = 5,
  confounders_baseline = c("sex", "education_years", "bmi", "smoke_binary")
)

res_atc <- estimate_yll_gformula_atc(
  data = yll_toy,
  B = 50, method = "normal", show_progress = FALSE, use_future = FALSE,
  id_var = "id", time_var = "period", event_var = "event",
  exposure_var = "hypertension",
  reference_level = "No", exposed_level = "Yes",
  age_at_entry_var = "age",
  age_start = 50, age_end = 90, age_interval = 5,
  confounders_baseline = c("sex", "education_years", "bmi", "smoke_binary")
)
```

## Stochastic interventions

For policy-relevant questions where "set everyone to exposed" is
unrealistic, use a **stochastic intervention** that shifts the exposure
distribution. The convenience wrapper
`estimate_yll_gformula_binary_stochastic_vs_natural()` compares a target
intervention to the natural exposure distribution observed in the data.

The example below contrasts a world in which 20% of the unexposed and
100% of the exposed end up exposed against the natural distribution:

```{r stochastic, eval = FALSE}
res_increase <- estimate_yll_gformula_binary_stochastic_vs_natural(
  data = yll_toy,
  prob_exposed_if_unexposed = 0.2,
  prob_exposed_if_exposed   = 1.0,
  target_population = NULL,
  B = 50, show_progress = FALSE, use_future = FALSE,
  id_var = "id", time_var = "period", event_var = "event",
  exposure_var = "hypertension",
  reference_level = "No", exposed_level = "Yes",
  age_at_entry_var = "age",
  age_start = 50, age_end = 90, age_interval = 5,
  confounders_baseline = c("sex", "education_years", "bmi", "smoke_binary")
)
```

See `vignette("interventional-effects", package = "YLLgmethods")` for the
underlying theory and more examples.

## Visualisation

Three plotting helpers are included; all return `ggplot2` objects so
you can post-process freely.

```{r plots, eval = FALSE}
plot_yll(res_ate)
plot_conditional_survival(res_ate, age_start = 60)
plot_marginal_survival(res_ate)
```

## Reproducibility tips

- Set `seed` to make the bootstrap deterministic.
- Set `use_future = TRUE` and configure a `future::plan()` to
  parallelise the bootstrap.
- `integration = "trapezoidal"` produces slightly smoother LE estimates;
  the default `"left_rectangle"` is consistent with the discrete-time
  hazard model.
