---
title: "Generalized interventional effects with YLLgmethods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized interventional effects with YLLgmethods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how `YLLgmethods` represents general
**interventional effects** — including stochastic and dynamic regimes —
and how that maps onto the package API.

## 1. From static interventions to a more general framework

The classical average treatment effect (ATE) compares two **static**
worlds, e.g. *everyone non-smoker* vs. *everyone current smoker*:

\[
\mathrm{ATE} = E[Y^{a=1}] - E[Y^{a=0}].
\]

This is conceptually clean but often unrealistic. **Generalized
interventional effects** instead define an effect over a *class* of
interventions — typically by specifying a stochastic assignment
distribution \(g(a \mid L)\) and contrasting two such distributions:

\[
E[Y^{g_1}] - E[Y^{g_0}].
\]

The g-formula identifies these as

\[
E[Y^g] = E_L\!\left[\, \sum_a E[Y \mid A = a, L]\, g(a \mid L) \right].
\]

## 2. What changes for binary exposures

Suppose the observed exposure is binary,
\(A \in \{\text{reference}, \text{exposed}\}\), and we want to contrast:

- **Target intervention.** "Among those who were observed unexposed,
  20% become exposed; among those observed exposed, all stay exposed."
- **Reference intervention.** Some other shift (or the natural exposure
  distribution).

For each subject the g-formula needs only \(P(A^* = \text{exposed} \mid A, L)\)
under the two regimes. `YLLgmethods` exposes this through:

- `yll_make_binary_stochastic_intervention()` — builds a function
  `f(data)` returning subject-specific exposure probabilities, indexed
  by the subject's *originally observed* exposure status.
- `estimate_yll_gformula_intervention()` — takes
  `intervention_reference` and `intervention_exposed`, each either a
  scalar probability, a vector of probabilities, an exposure-level
  label, or a function built with the helper above.
- `estimate_yll_gformula_binary_stochastic()` and
  `..._binary_stochastic_vs_natural()` — convenience wrappers around the
  most common binary-exposure shifts.

## 3. Worked example: shifting hypertension prevalence

Imagine a public-health programme that would, if applied uniformly to
the unexposed group, expose 20% of them and keep all observed-exposed
subjects exposed. We compare it to the **natural** exposure distribution
that produced the observed data.

```{r demo, eval = FALSE}
library(YLLgmethods)
data(yll_toy)

res <- estimate_yll_gformula_binary_stochastic_vs_natural(
  data = yll_toy,
  prob_exposed_if_unexposed = 0.20,
  prob_exposed_if_exposed   = 1.00,
  target_population = NULL,        # average over the entire population
  B = 50, show_progress = FALSE, use_future = FALSE,
  id_var = "id", time_var = "period", event_var = "event",
  exposure_var = "hypertension",
  reference_level = "No", exposed_level = "Yes",
  age_at_entry_var = "age",
  age_start = 50, age_end = 90, age_interval = 5,
  confounders_baseline = c("sex", "education_years", "bmi", "smoke_binary")
)

res$summary
```

The "natural" intervention reuses the empirical exposure assignments,
so subjects observed as `Yes` get \(P(A^* = \text{Yes}) = 1\) and those
observed as `No` get \(0\). Internally this is just a degenerate
stochastic intervention — exactly the same machinery.

## 4. Custom regimes

For arbitrary regimes (covariate-dependent assignment, dynamic rules,
shift-based interventions, …), pass a function directly to
`estimate_yll_gformula_intervention()`:

```{r custom, eval = FALSE}
intervention_target <- function(data) {
  # Higher exposure probability for older subjects
  plogis(-1 + 0.05 * (data$age - 55))
}

res <- estimate_yll_gformula_intervention(
  data = yll_toy,
  intervention_reference = "No",        # natural reference: nobody exposed
  intervention_exposed   = intervention_target,
  id_var = "id", time_var = "period", event_var = "event",
  exposure_var = "hypertension",
  reference_level = "No", exposed_level = "Yes",
  age_at_entry_var = "age",
  age_start = 50, age_end = 90, age_interval = 5,
  confounders_baseline = c("sex", "education_years", "bmi", "smoke_binary"),
  B = 0, use_future = FALSE
)
```

The `intervention_*` arguments accept:

- a scalar probability in `[0, 1]`,
- a vector of length `nrow(data)` of probabilities,
- a value matching one of the observed exposure levels (treated as a
  degenerate intervention), or
- a function `f(data)` returning either of the above.

## 5. Identification assumptions

For these estimands to recover the true counterfactual contrast, the
usual g-formula conditions are required:

- **Consistency** of the counterfactual outcomes under the
  hypothesised regimes.
- **Conditional exchangeability** \(Y^a \perp A \mid L\).
- **Positivity** of the shifted exposure regimes given the chosen
  conditioning set.
- A correctly specified outcome model — here a discrete-time pooled
  logistic hazard with natural cubic splines on attained age.

Stochastic interventions are particularly attractive when fully static
regimes are infeasible or violate positivity, because the shifts can be
designed to stay within the empirical support of the data.
